---
title: "Merge and Analyze"
author: "Zachary Winn"
date: "3/5/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# MLMs

## Intro and Data Read-In

This is a quick r script I wrote to merge the phenotype files and then analyze.

```{r merge data}
library(ggplot2)
library(tidyverse)
library(asreml)
library(psych)
library(knitr)
library(GGally)

setwd("C:/Users/zwinn/OneDrive/University of Arkansas/Post Graduation Work/Analysis")

hgawn<-read.csv("HGAWN_Aligned_Data.csv")

hgawn<-hgawn %>% 
  filter(!Variety=="6-Row-Barley") %>%
  select(Tray, Rep, Variety, 
         Days_to_heading, Date_of_anther_sampling, 
         Area_of_anthers_per_spike, Number_of_anthers_per_spike,
         Area_per_anther)

colnames(hgawn)[4:5]<-c("Heading_Date", "Anthesis_Date")

hgawn[,1:3]<-lapply(hgawn[,1:3], as.factor)
hgawn[,4:ncol(hgawn)]<-lapply(hgawn[,4:ncol(hgawn)], as.numeric)
lapply(hgawn[,1:ncol(hgawn)], class)

head(hgawn)
```

## Linear Models

Here I loop through the to perform the following mixed linear model in ASReml:  

$y=V+R+\varepsilon$ 

Where y is the response, V is the variety which is fixed, R is the rep which is the blocking factor and random, an $\varepsilon$ is the residual error. 

```{r linear models}
#set up traits
traits<-colnames(hgawn[4:ncol(hgawn)])

#create dataframe to bind onto
preds<-data.frame(Variety=unique(hgawn$Variety))

#create dataframe for AIC, BIC, and LogLik
crit<-data.frame(row.names = c("AIC", "BIC", "LogLik"))

#create table for paper
table<-data.frame(row.names = c("Trait", "DF", "Sum of Squares",
                                "Wald statistic", "P-Value", "Repeatability",
                                "Standard Error"))

table<-as.data.frame(t(table))

#look at distribution
for (i in traits){
  hist(hgawn[,i], main = paste("Distribution of", i), xlab = i)
  boxplot(hgawn[,i], main=i)
}

#loop MLMs for each trait and pull BLUEs
for (i in traits){
  #announce step
  print(paste("------- Analyzing trait", i, "-------"))
  
  #run model
  mlm<-asreml(fixed=hgawn[,i]~1+Variety,
              random=~Rep,
              units=~idv(units),
              data=hgawn)
  
  #display anova table
  a<-wald.asreml(mlm)
  print(a)
  
  #look at residuals
  print(plot(resid(mlm), main=paste("Residuals of", i, "Analysis")))
  
  #pull out fit criteria
  c<-as.data.frame(rbind(summary.asreml(mlm)$aic, summary.asreml(mlm)$bic, summary.asreml(mlm)$loglik))
  colnames(c)<-i
  crit<-cbind(crit,c)
  
  #Pull out fixed effects
  p<-predict(mlm, classify = "Variety")$pvals[,1:2]
  colnames(p)[2]<-i
  preds<-left_join(preds, p)
  
  #run a random model to pull out estimates of repeatability
  mlm<-asreml(fixed=hgawn[,i]~1,
              random=~Variety+Rep,
              units=~idv(units),
              data=hgawn)  
  
  #predict repeatability
  r<-vpredict(mlm, r~V2/(V1+V2+V3))
  
  t<-cbind(i,as.data.frame(a)[2,],r)
  colnames(t)[1]="Trait"
  
  table<-rbind(table,t)
  
  #remove
  remove(mlm,a,r,c,p)
}

#output table
write.csv(table, "Table 1.csv", row.names = F)

#look at the fit criteria
kable(crit, caption="Fit Criteria of RCBD for Each Trait")

#write out predictions
write.csv(preds,"BLUEs_2018_2019_HGAWN.csv", row.names = F)

#remove outlier for correlation study
preds<-preds[-as.numeric(rownames(preds)[which.max(preds$Area_of_anthers_per_spike)]),]

#make a pretty figure with no underscore
fig<-preds
colnames(fig)<-c("Variety","HD","AD","AOAPS","NOAPS","APA")

#display the raw values of the correlation
kable(cor(fig[,2:ncol(fig)], use="complete.obs"), caption="Correlations Among BLUEs")


#figure for publication
a<-ggpairs(fig[,2:ncol(fig)],
        lower = list(continous=wrap("smooth", color="blue")),
        diag =  list(continous="barDiag"),
        upper =  list(continous="cor")) 

a+theme(legend.position = "none", 
        panel.grid.major = element_blank(), 
        axis.ticks.x = element_blank(), 
        panel.border = element_rect(linetype = "dashed", colour = "black", fill = NA))

library(GGally)

ggpairs(fig[,2:ncol(fig)])
pairs.panels(fig[,2:ncol(fig)], stars = TRUE)
```

## PCA Analysis

```{r}
#create a correlation matrix
S<-cor(fig[,2:ncol(fig)], use = "complete.obs")

#perform eigen value decomposition on the correlation matrix
V<-eigen(S)$vectors
V<-as.data.frame(V, row.names = rownames(S))
V<-rownames_to_column(V, var="Trait")
colnames(V)<-paste0("PC", 1:5)
R<-eigen(S)
R<-as.data.frame(R$values/5)

#change point size
update_geom_defaults("point",list(size=4))

#2D plot loadings
ggplot(data = V, aes(x=PC1, y=PC2, color=Trait))+
  geom_point()+
  theme(legend.title=element_blank())

#library
library(plotly)

#3D plot of loadings
plot_ly(data=V, x=~PC1, y=~PC2, z=~PC3, color=~Trait) %>% add_markers()
```